Retirement Consumption: Evidence from UK Pension Reform

### 2023-07-26 Texfiles/Draft1.tex
latest edits

\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}
\setlength{\parindent}{0pt}
\usepackage[parfill]{parskip}
\date{Lent 2023}
\title{Dissertation draft}
\author{Tobias Leigh-Wood}
\usepackage[letterpaper, margin=1.1in]{geometry}
\linespread{1.1}

@@

\maketitle
\tableofcontents
\newpage

Standard economic theory suggests that retirement annuities, which provide a guaranteed income stream until the end of life,
should be highly valued by individuals as a way to insure against late death \cite{yaari_65}. However, in developed countries,
rates of annuitisation are far below the levels that theory predicts. The literature has suggested several reasons for
this, but there is no consensus about which mechanism is dominant. I exploit early retirees' consumption responses to a
reform to annuity policy in the UK to add to this literature.
    [\textbf{Add something once I have a conclusion}].

In the UK, adults have three methods of funding retirement: the state pension; defined benefit (DB) employee pension schemes,
which provide an income in retirement based on years of service and a function of wages; and defined contribution (DC) pensions,
in which individuals save and invest in a tax-advantaged account that is usually supplemented with employer contributions.
Under the 2010-2015 coalition government in the UK, the law regarding the use of private DC pensions at
retirement changed. Individuals were no longer forced to annuitise, which provides a constant income until death,
their pension pots and could access them in a variety of ways such as a lump sum withdrawal or income drawdown,
which is a steady withdrawal of assets
with the remainder left in the pot - common advice is to take 4\% a year.
Subsequently the number of annuities sold in the UK dropped precipitously.

In this paper, I first use the policy reform as a discontinuity to measure the impact of forced annuitisation on the consumption
levels of individuals in the first few years of retirement. Given that the reform was implemented suddenly and without
advanced notice before the Spring 2014 budget, I claim that individuals who retired in 2015, 2016 or 2017
are otherwise similar to those who retired year in 2011, 2012 and 2013 except for the fact the later retirees
do not need to annuitise their DC pension pots. Therefore, I can use a regression discontinuity to identify
the effect that forced annuitisation has on the consumption of retirees.

I then test two competing hypotheses for the annuity puzzle: bequests and
pessimistic life expectancy. The annuity puzzle was first identified by \cite{yaari_65} and is a classed as a problem
since standard economic theory cannot explain the lack of annuitisation that is observed in the real world.
Two possible explanations are bequests and life expectancy: individuals want to leave an estate for their heirs to
inherit and they cannot do this if they annuitise their wealth; or individuals may believe they will not live
as long as annuity providers think they will and therefore annuities that are on the market do not appear to be
good value. Depending on which reason explains the lack of annuitisation in the UK, the consumption response of
retirees to the pension
reform will differ. If individuals do not annuitise because of pessimistic life expectancy their
consumption should increase after the reform. If, on the other hand, individuals do not annuitise because of a bequest motive, consumption
should not change much as a result of the reform.

@@

The importance of retirement policy to individuals in the UK is growing. The number of individuals of
pensionable age is expected to increase from 11.9 million in 2020 to 15.2 million in
2045 according to the latest Office for National Statistics (ONS) and for every 1000 people of working age there will be 341 of pensionable
age in 2045 compared to 280 in 2020 \cite{ons_population_predictions_2020}. The increase in absolute and relative
numbers of elderly people makes retirement policy more consequential. Moreover, private DC pensions are
becoming increasingly common and are predicted to grow as current cohorts age \cite{cribb_karjalainen_ifs_2023}.
Therefore, policies regarding how private pensions can be accessed will have a larger impact on overall welfare for
retirees.

Moreover, understanding how retirees spend their money over retirement is an important policy question of its own.
If retirees spend too much in early retirement, the state may need to provide for them towards the end of their
lives. This has implications for government budgets, especially since population aging means more individuals will
require expensive end of life care. On the other hand, if retirees over-save and do not spend, the economy may face
a dynamic inefficiency whereby consumption per capita could be increased if savings were decreased. (Get dynamic inefficiency
reference.)

\subsection{Literature review}
My paper builds on three main strands of literature: that of the annuity puzzle, exploring reasons that some retirees choose not to annuitise;
the retirement saving puzzle, looking at consumption and saving behaviour over retirement; and lifecycle models, whereby the above
problems are explained using a tractable model of human decision making.

\cite{yaari_65} was the first to show that under standard assumptions we would expect individuals to
annuitise all of their wealth at retirement to insure against the risk of long life. Since then there
has been much literature discussing possible reasons that people do not annuitise. \cite{finkelstein_porteba_2002}
and \cite{finkelstein_porteba_2004} find evidence of adverse selection, thereby making the 'money's worth'
of annuities lower for the general population as opposed to the population of annuitants. However, they
also find that theory would still predict annuitisation.

@@

\textbf{If I add 300 more words here then I will be at 5000!}

@@

The main data set I use is the English Longitudinal Study of Ageing (ELSA) \cite{main_elsa_citation}. ELSA picks individuals
over the age of 50 to survey every two years until death. If individuals leave the survey ELSA replaces them so that it is
representative of the over 50 population in the UK. Individuals are asked a range of questions relating to their income
and wealth as well as expectations about the future. One benefit of using ELSA is that it includes data on pension types
for individuals who are working. Therefore, I can differentiate between individuals who have a DC pension and those who have a DB pension.
There have been 9 waves of ELSA data collection, the first was in 2001 and collections happened every two years thereafter.

Importantly ELSA also includes information on pension size that the Institute for Fiscal Studies (IFS) calculated, this is only available
up until wave 5 (which was collected in 2010 and 2011) at which point I use a real return
of 3\% to predict forward the pension wealth variable until retirement. ELSA also includes a measure of all
non-housing financial wealth which I use in some specifications because of these issues with
pension wealth. Due to slight differences in the ELSA questions between years, I use `Harmonized ELSA'
\footnote{This analysis uses data or information from the Harmonized ELSA dataset and Codebook, Version G.2 as of
    July 2021 developed by the Gateway to Global Aging Data. The development of the Harmonized ELSA was funded by the National
    Institute on Aging (R01 AG030153, RC2 AG036619, R03 AG043052). For more information,
    please refer to https://g2aging.org/.} which ensures that variables are comparable across waves. Since this only includes a
subset of the questions in ELSA, I also supplement it with variables taken directly from the data such as questions that deal
with life expectancy.

ELSA also includes questions on expenditures. In particular, individuals are asked how much they consume across a range of broad
categories including in-house food consumption, out-of-house food consumption, leisure consumption, clothes consumption and
consumption on utility bills and rent. I use also use two variables related to life expectancy that I discuss in more detail below.

I also use `life tables' from the UK's ONS. These provide me with risk of death for each age group. I adjust them
to make death certain at age 110 as is common in the literature. I transform these so that I have risk of death conditional on being
a given age since this is what I use in the life cycle simulations. I also use these objective probabilities to
calculate annuity prices for individuals.

To illustrate the effect the reform had on sales of annuities in the UK I obtained product data from the Financial Conduct Authority.
These track the sales of different financial products overtime including data annuities.

Life expectancy impacts decision to retire for two reasons. Firstly, it directly impacts the price that an annuity will cost for
individuals. Older individuals and those with pre-existing health conditions generally can buy an annuity that provides a greater
income stream than individuals who are younger. Secondly, private information about life expectancy impacts the perceived
value of an annuity. If an individual expects to outlive the general population an annuity, at a given price, will appear
a much better deal to them. Likewise, an individuals life expectancy will change their decision on how to consume and save
through retirement.

To calculate subjective life expectancies I follow \cite{odea_sturrock_rest_2023}. Individuals are asked “What are the chances that you will
live to be age X or more?” where X changes depending on the age of the interviewee. If individuals were under 65 then X was 75, if individuals
were 66 and older they were asked the age that was 11 to 15 years older than them and is a multiple of 5. From wave three respondents were
also asked “What are the chances that you will live to be age 85 or more?” if they were under 70. As most recent retirees are under 70 we
therefore have two data points. I first drop from the data individuals who think it is more likely that they reach a higher age than a younger
age since this shows a misunderstanding of the question. I then add as a third data point their objective chance of reaching 110 according to the ONS life tables.
I fit these three points to a Weibull distribution, which is commonly used by demographers to estimate how populations
will age, using non-linear least squares. Then I create subjective survival tables using parameters from the Weibull distribution.

@@

Successive governments in the UK have attempted to reform both the public and private pension system. Prior
to 1987, participation in private schemes was limited to employees of firms that had offered them, and there
were few alternatives to the public state pension or defined benefit scheme that a public sector employer
would offer. From 1989, individuals in the UK were allowed to open tax advantaged self-invested personal pensions
alongside any pension their employer offered. The 2004 Finance Act rationalised taxation rates on different types of pensions.
But DC pension pots could still only be used to buy an annuity after taking out a maximum of 25\% as a tax free lump sum withdrawal
from age 55 onwards. Most pension pots also required that individuals access them before age 75.

Announced in the spring budget of 2014, the so called `pensions freedom act' received Royal Asset in December 2014 marked the end of a series
of pension reforms carried out by the coalition government between 2010 and 2015. The reform made it possible
to withdraw the first 25\% from a pension pot tax free and the rest subject to the marginal rate of income tax that an individual faced.
In the June 2010 budget, the government made a first reform to the annuitisation rules, creating
a minimum income requirement above which individuals would not need to
annuitise more \cite{finance_act_hmt_2011} (this meant high income retirees would not need to annuitise their wealth).
However, this was set at £20,000 and therefore few individuals
were eligible. The minimum income requirement was scrapped in the 2014 bill, finally eliminating the
compulsory annuities market, with one government minister saying pension pots `can be used to buy Lamborghinis'
\cite{guardian_lambos}.

The impact of the reforms on annuity demand has been documented by \cite{cannon_et_al_nier_2016}.
Using data from the Association of British Insurers, they show that annuity demand dropped by 75\% from
its maximum in 2011
\textbf{Check}
. Figure \ref{fig:annovertime} shows purchases of annuities overtime and demonstrates the sharp decrease
in purchases that happened from 2014 to 2015. There was an increase in the number of pensions that were
being accessed using an income drawdown product, but these do not fully account for the drop in annuitisation.
\begin{figure}[h]
    \caption{Pension pots accessed }
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/annuity_overtime.pdf}
    \label{fig:annovertime}

\end{figure}

\begin{figure}[h]
    \caption[Caption for LOF]{How pension pots are accessed\protect\footnotemark}
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/annuity_pot_sizes.pdf}
    \label{fig:ann2122}
    \
\end{figure}
Figure \ref{fig:ann2122} shows the distribution of how pension pots were first accessed at retirement in 2021-22.
In 2021, 196,736 pots were fully withdrawn at retirement, accounting for over 50\% of pots. Prior to the policy
reform this was not the case -- most defined contribution pensions were accessed through annuities.
\footnotetext{Partial UFPLS (uncrystallised funds pension lump sum) is similar to Partial drawdown but with a different tax schedule. With partial drawdown
    you take the whole 25\% tax free amount at once whereas with UFPLS you only claim the 25\% on the amount you are taking out. This option is
    better if you may want to buy another retirement product in the future such as an annuity. }

Also of specific interest to the annuity market was the European Union's
`Equal Treatment in Goods and Services Directive' of 2004. This prohibited discrimination
in the provision and cost of goods and services based on sex. Up until 2011, there was a clause that
stated insurers were allowed to charge different premiums if it was based on evidence that sex is
correlated with different amounts of risk. However, in March 2011, the European Court of Justice
ruled that insurers were not allowed to charge different amounts and gave them until December 2012
to implement the ruling. This change meant that annuity products could no longer be priced differently
for men and women in the UK. However, given that this change was implemented two years before the pension
freedoms act, I can still identify the impact of the pensions freedom act on early retirement consumption.

@@

As required, retirement year is between 2011 and 2013 for the control group and 2015 and 2017 for the treatment group.
However, as discussed above their interview date is after or the same as the retirement year because we are tracking
consumption in retirement.

The second retirement group, those who retired post reform, is small with just 728 non missing observations for gender as opposed
to 941 individuals in the control group. The control group are more female, retired at a slightly younger age and expected to
retire slightly earlier. This could be because this period was affected by the increase in the state pension age for
women from age 60 in 2010 to age 65 in 2018. Since this change happened gradually and occured over the whole period
I do not expect it to influence the results. The DifferenceAge row tracks the difference between expected and actual
pension age.

The treatment group has higher financial wealth with a median of £66,000 at time of interview as opposed to £55,000
in the control group. Likewise, the second group are more likely to have held a DC pension at some point
and have much more money in them. Both groups are equally likely to have a defined benefit pension with roughly 44\% of
individuals across both samples having a defined benefit pension. In general, DB pensions are more prevalent than
DC pensions in the data, this reflects the same trend that is observed in population of UK retiree. Home ownership
is roughly equal across groups although the second group have slightly higher housing wealth.
\textbf{I have
    not inflation adjusted but I am thinking about it?}

Both groups are similarly long-lived according when using ONS life tables to calculate life expectancy
based on gender, age and the year the interview was carried out in. Subjective life expectancies are also similar
across groups, with individuals expecting to live another 21 years as opposed to the 24 that the ONS would expect
them to live.
\textbf{Maybe add in prob of survival to age 85 or 90 or something}
\textbf{Also add in an indicator for living alone or living with someone else}

@@

\textbf{Check why total consumption adds up to so much
    more than all the categories added together? Do they include rent and housing in consumption?}

Overall, the groups appear to have slightly different wealth profiles with the treatment group being richer and slightly more likely
to have DC pension wealth. On key demographic characteristics, however, the groups are similar, the difference in retirement
age between the two groups can be mostly explained by the difference in expected retirement age meaning that individuals
have not en masse decided to delay retirement and avoid annuitisation.

\begin{landscape}

    \input{tables/r_summary_stats.tex}

\end{landscape}

As a further test for balance I regress demographic and financial characteristics of individuals on year of birth and the treatment dummy.
We can then see if the treatment groups differ on key characteristics such as financial wealth or retirement age.
For a regression discontinuity to be valid we need the treatment and control groups to be similar along
all other characteristics apart from the treatment variable.

In particular I run:
\begin{equation*}
    Y_{i} = \alpha + \beta Post_{i} + \gamma YOB_{i} + \kappa (Post_{i} \times YOB_{i}) + \epsilon_{i}
\end{equation*}

\input{tables/cov_balance.tex}
Table \ref{tab:cov_balance} shows these results. We can see that being in the treatment group.
\textbf{I am still thinking about this.}

One threat to validity is manipulation of the running variable -- which is the variable that decides
whether an individual is in the treatment group or not -- in our case retirement year.
This is probably quite a threat in this setting as individuals could delay retirement and/or
the purchase of an annuity until after the reform. To check whether
this is happening we see whether the difference between real retirement age and the first expected retirement age
recorded in the data are different for those retiring before 2014 compared to those retiring after 2014. As shown
above in Table \ref{tab:sum_stats} this not the case.

However, it is possible that individuals do not buy an annuity at the time of retirement since the law prior to 2014 only
required that they access the pot by age 75. They may decide to keep their DC pension untouched
and live off other income and assets before accessing it later. Since we do not have data on the
exact purchase date of annuities we cannot track whether
annuity purchases were delayed because of this. However, there is relatively stable annuity demand
before the policy reform. If individuals delayed annuity purchases in expectation of the reform we
would observe declines in purchases before the reform was announced.

@@

In this section I outline the key empirical models I run with both the simulated consumption data from the
lifecycle models and with the real data from ELSA. I then see which lifecycle model better fits the consumption
response that occurred as a result of the pension reform.

I use a regression discontinuity design for which I compare the consumption of early retirees after the policy reform
to that of people who retired just before the policy reform. The key assumption implicit in regression discontinuities is
that nothing else changes at the time of the jump apart from the policy of interest, and that the policy occurs
without individuals predicting it and altering their behaviour. As I have argued above, the demographic information in the
data suggest that individuals did not delay retirement and that there was no delay in annuitisation as shown by the quick
and sudden decline in annuity purchases. Moreover, anecdotal evidence from media and business sources at the time of the
announcement show that there was surprise the government had gone this far, with money marketing describing the change as a
"bombshell" \cite{money_marketing_announcement}.

Retirement year is the running variable and individuals are treated if retirement year is greater than 2014 and less than
or equal to 2017. I use consumption data of individuals up to 2 years into retirement so that the sample size is larger.
So if someone retired in 2015 and had consumption data in 2015 and 2017, I include both values. An individual is considered
not treated if they retire before 2013 and after 2011.  To account for the fact
that the reform only impacted individuals who had accumulated defined contribution pension pots I interact the treatment dummy with
an indicator variable signalling whether the individual had ever held a DC pension pot.

Because of the differences in financial characteristics between the groups, I add controls for financial wealth, housing wealth
and whether an individual owns their own home.

@@

And $X_{it}$ is a set of controls.

>>>>

\hypersetup{ colorlinks, citecolor=black, filecolor=black, linkcolor=black,
    urlcolor=black }
\setlength{\parindent}{0pt}
\usepackage[parfill]{parskip}
\date{July 2023}
\title{Retirement Consumption: Evidence from UK Pension Reform}
\author{INSERT NUMBER HERE}
\usepackage[letterpaper, margin=1.1in]{geometry}
\linespread{1.25}

@@

\newpage
\maketitle
\tableofcontents
\newpage

Classical economic theory suggests that retirement annuities, which provide a
guaranteed income stream until the end of life, should be highly valued by
individuals as a way to insure against late death \cite{yaari_65}. However, in
developed countries, rates of annuitisation are far below the levels that theory
predicts - this phenomenon has been termed the "annuity puzzle". The literature
has suggested several reasons for this, but there is no consensus about which
mechanism is dominant. I exploit early retirees' consumption responses to a
reform to annuity policy in the UK to add to this literature. [\textbf{Add
            something once I have a conclusion}].

In the UK, adults have three methods of funding retirement: the state pension;
defined benefit (DB) employee pension schemes, which provide an income in
retirement based on years of service and a function of wages; and defined
contribution (DC) pensions, under which individuals save and invest in a
tax-advantaged account that is usually supplemented with employer contributions.
Under the 2010-2015 coalition government in the UK, the law regarding the use of
private DC pensions at retirement changed. Individuals were no longer forced to
annuitise their pension pots and could access them in a variety of ways, such as
a lump sum withdrawal or income drawdown, involving a steady withdrawal of
assets from the pension pot -- common advice is to take 4\% a year. Subsequently
the number of annuities sold in the UK dropped precipitously.

In this paper, I first use the policy reform as a discontinuity to measure the
impact of forced annuitisation on the consumption levels of individuals in the
first few years of retirement. Given that the reform was implemented suddenly
and without advanced notice before the Spring 2014 budget, I claim that
individuals who retired in 2015, 2016 or 2017 are otherwise similar to those who
retired in 2011, 2012 and 2013 except for the fact the later retirees do not
need to annuitise their DC pension pots. Therefore, I can use a regression
discontinuity to identify the effect that forced annuitisation has on the
consumption of retirees.

I then test two competing hypotheses for the annuity puzzle: bequests and
pessimistic life expectancy. The annuity puzzle was first identified by
\cite{yaari_65} and is a classed as a problem since standard economic theory
cannot explain the lack of annuitisation that is observed in the real world. Two
possible explanations are bequests and life expectancy: individuals want to
leave an estate for their heirs to inherit and they cannot do this if they
annuitise their wealth; or individuals may believe they will not live as long as
annuity providers think they will and therefore annuities that are on the market
do not appear to be good value. Depending on the reason for the lack of
annuitisation in the UK, the consumption response of retirees to the pension
reform will differ. If individuals do not annuitise because of pessimistic life
expectancy their consumption should increase after the reform. If, on the other
hand, individuals do not annuitise \ because of a bequest motive, consumption
should not change much as a result of the reform.

@@

The importance of retirement policy to individuals in the UK is growing. The
number of individuals of pensionable age is expected to increase from 11.9
million in 2020 to 15.2 million in 2045 according to the latest projections from
the Office for National Statistics (ONS) and for every 1000 people of working
age there will be 341 of pensionable age in 2045 compared to 280 in 2020
\cite{ons_population_predictions_2020}. The increase in absolute and relative
numbers of elderly people makes retirement policy more consequential. Moreover,
private DC pensions are becoming increasingly common and are predicted to grow
as current cohorts age \cite{cribb_karjalainen_ifs_2023}. Therefore, policies
regarding how private pensions can be accessed will have a larger impact on
overall welfare for retirees.

Moreover, understanding how retirees spend their money over retirement is an
important policy question of its own. If retirees spend too much in early
retirement, the state may need to provide for them towards the end of their
lives. This has implications for government budgets, especially since population
aging means more individuals will require expensive end of life care. On the
other hand, if retirees over-save and do not spend, the economy may be
dynamically inefficient whereby the capital stock is too high because
individuals save too much for retirement and for bequests. In this case
consumption per capita could be increased if savings were decreased.

\subsection{Literature review}
My paper builds on three main strands of literature: that of the annuity puzzle,
exploring reasons that some retirees choose not to annuitise; the retirement
saving puzzle, which seeks to explain why retirees drawdown assets slowly, ; and
lifecycle models, whereby the above problems are explained using a tractable
model of human decision making.

\cite{yaari_65} showed that under standard assumptions we would expect
individuals to annuitise all of their wealth at retirement to insure against the
risk of long life. Since then there has been much literature discussing possible
reasons that people do not annuitise. \cite{finkelstein_porteba_2002} and
\cite{finkelstein_porteba_2004} find evidence of adverse selection, thereby
making the 'money's worth' of annuities lower for the general population as
opposed to the population of annuitants. However, they also find that theory
would still predict annuitisation.

@@

@@

The main dataset I use is the English Longitudinal Study of Ageing (ELSA)
\cite{main_elsa_citation}. ELSA picks individuals over the age of 50 to survey
every two years until death. If individuals leave the survey, ELSA replaces them
so that it is representative of the over 50 population in the UK. Individuals
are asked a range of questions relating to their income and wealth as well as
expectations about the future. One benefit of using ELSA is that it includes
data on pension types for individuals who are working. Therefore, I can
differentiate between individuals who have a DC pension and those who have a DB
pension. There have been 9 waves of ELSA data collection -- the first was in
2001 and collections happened every two years thereafter.

Importantly, ELSA also includes information on pension size calculated by the
Institute for Fiscal Studies (IFS). This is only available up to and including
wave 5 (which was collected in 2010 and 2011) at which point I use a real return
of 3\% to predict forward the pension wealth variable until retirement. ELSA
also includes a measure of all non-housing financial wealth which I use in some
specifications because of these issues with pension wealth data. Due to slight
differences in the ELSA questions between years, I use `Harmonized ELSA'
\footnote{This analysis uses data or information from the Harmonized ELSA
    dataset and Codebook, Version G.2 as of July 2021 developed by the Gateway to
    Global Aging Data. The development of the Harmonized ELSA was funded by the
    National Institute on Aging (R01 AG030153, RC2 AG036619, R03 AG043052). For more
    information, please refer to https://g2aging.org/.} which ensures that variables
are comparable across waves. Since this only includes a subset of the questions
in ELSA, I supplement it with variables taken directly from the data such as
questions that deal with life expectancy.

ELSA also includes questions on expenditures. In particular, individuals are
asked how much they consume across a range of broad categories including
in-house food consumption, out-of-house food consumption, leisure consumption,
clothes consumption and consumption on utility bills and rent.

I also use `life tables' from the UK's ONS. These provide me with the risk of
death for each age group. Since the tables are produced until age 120, I adjust
them to make death certain at age 110 as is common in the literature
\cite{odea_sturrock_rest_2023}. I transform these so that I have risk of death
conditional on being a given age since this is what I use in the life cycle
simulations. I also use these objective probabilities to calculate annuity
prices for individuals.

To illustrate the effect that the reform had on sales of annuities in the UK I
obtained product data from the Financial Conduct Authority. These track the
sales of different financial products overtime including data on annuities.

Life expectancy impacts the decision to annuitise for two reasons. Firstly, it
directly impacts the price that an annuity will cost for individuals. Older
individuals and those with pre-existing health conditions generally can buy an
annuity that provides a greater income stream than individuals who are younger.
Secondly, private information about life expectancy impacts the perceived value
of an annuity. If an individual expects to outlive the general population, an
annuity, at a given price, will appear a much better deal to them. Likewise, an
individual's life expectancy will change their decision on how to consume and
save through retirement.

To calculate subjective life expectancies from ELSA data, I follow
\cite{odea_sturrock_rest_2023}. Individuals were asked “What are the chances
that you will live to be age X or more?” where X changed depending on the age of
the interviewee. If individuals were under 65 then X was 75, if individuals were
66 and older they were asked the age that was 11 to 15 years older than them and
a multiple of 5. From wave 3 respondents were also asked “What are the chances
that you will live to be age 85 or more?” if they were under 70. As most recent
retirees are under 70 we therefore have two data points. I first drop from the
data individuals who think it is more likely that they reach a higher age than a
younger age since this shows a misunderstanding of the question. I then add as a
third data point their objective chance of reaching 110 according to the ONS
life tables. I fit these three points to a Weibull distribution, which is
commonly used by demographers to estimate how populations will age, using
non-linear least squares. Then I create subjective survival tables using
parameters from the Weibull distribution.

@@

Successive governments in the UK have attempted to reform both the public and
private pension system. Prior to 1987, participation in private schemes was
limited to employees of firms that had offered them, and there were few
alternatives to the public state pension or DB scheme that a public sector
employer would offer. From 1989, individuals in the UK were allowed to open tax
advantaged self-invested personal pensions alongside any pension their employer
offered. The 2004 Finance Act rationalised taxation rates on DB and DC pensions.
The rules for DC pension pots were such that individuals had to buy an annuity
after optionally taking a maximum of 25\% of the pot as a tax free lump sum
withdrawal. DC pension pots were accessible from age 55 and most required that
they be accessed before the age of 75.

In the June 2010 budget, the coalition government made the first of two key
reforms it would make to the pension system between 2010 and 2015. The 2010
policy reform created a minimum income requirement above which individuals would
not need to annuitise more \cite{finance_act_hmt_2011}, this meant high income
retirees would not need to annuitise their wealth. The income requirement was
set at £20,000. This meant, for example, that an individual who was a member of
a DB scheme paying them £20,000 a year would not need to annuitise their DC
pension pot. However, the relatively high minimum income requirement meant that
most individuals still had to purchase an annuity.

In spring 2014, George Osbourne announced the so called `pensions freedom act'
scrapping the minimum income requirement and eliminating the compulsory
annuities market \cite{pen_freedoms_hmt_2014}. One government minister
infamously brought the message home by saying pension pots `can be used to buy
Lamborghinis' if retirees wanted to \cite{guardian_lambos}. The government also
announced plans to make switches between DB and DC pension schemes possible
although this was only realised in the next parliament -- future research could
explore the impact of this second reform on the consumption and saving decisions
of retirees.

The impact of the reforms on annuity demand has been documented by
\cite{cannon_et_al_nier_2016}. Using data from the Association of British
Insurers, they show that annuity demand dropped by 75\% from its maximum in
2011. Figure \ref{fig:annovertime} shows purchases of annuities overtime and
demonstrates the sharp decrease in purchases that happened from 2014 to 2015.
There was an increase in the number of pensions that were being accessed using
an income drawdown product, but these only partly account for the drop in
annuitisation as is shown by the graph.
\begin{figure}[h]
    \caption{New pension pots accessed }
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/annuity_overtime.pdf}
    \label{fig:annovertime}
\end{figure}

\begin{figure}[h]
    \caption[Caption for LOF]{How pension pots are accessed\protect\footnotemark}
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/annuity_pot_sizes.pdf}
    \label{fig:ann2122}
    \
\end{figure}
Figure \ref{fig:ann2122} shows the distribution methods to access pension pots
at retirement in 2021-22. In 2021, 196,736 pots were fully withdrawn at
retirement, accounting for over 50\% of pots. Prior to the policy reform this
was not the case -- most defined contribution pensions were accessed through
annuities.

Also of specific interest to the annuity market was the European Union's `Equal
Treatment in Goods and Services Directive' of 2004. This prohibited
discrimination in the provision and cost of goods and services based on sex. Up
until 2011, there was a clause that stated insurers were allowed to charge
different premiums if it was based on evidence that sex is correlated with
different amounts of risk. However, in March 2011, the European Court of Justice
ruled that insurers were not allowed to charge different amounts and gave them
until December 2012 to implement the ruling. This change meant that annuity
products could no longer be priced differently for men and women in the UK.
However, given that this change was implemented two years before the pension
freedoms act it will not impact my results.

\footnotetext{Partial UFPLS (uncrystallised funds pension lump sum) is similar
    to Partial drawdown but with a different tax schedule. With partial drawdown
    you take the whole 25\% tax free amount at once whereas with UFPLS you only
    claim the tax relief on 25\% of the amount you are taking out. This option
    is better if you may want to buy another retirement product in the future
    such as an annuity. }

@@

As required, retirement year is between 2011 and 2013 for the control group and
2015 and 2017 for the treatment group. However, as discussed above their ELSA
interview date is after or the same as the retirement year because we are
tracking consumption in retirement.

The second retirement group, those who retired post-reform, is small with just
728 non missing observations for gender as opposed to 941 individuals in the
control group. The control group are more female, retired at a slightly younger
age and expected to retire slightly earlier. This could be because this period
was affected by the increase in the state pension age for women from age 60 in
2010 to age 65 in 2018. Since this change happened gradually and ocurred over
the whole period I do not expect it to influence the results. The DifferenceAge
row tracks the difference between expected and actual retirement age.

The treatment group has higher financial wealth with a median of £66,000 at the
time of interview as opposed to £55,000 in the control group. Likewise, the
second group are more likely to have held a DC pension at some point and have
much more money in them. Both groups are equally likely to have a DB pension
with roughly 44\% of individuals across both samples having a DB pension. In
general, DB pensions are more prevalent than DC pensions in the data, this
reflects the same trend that is observed in the general UK population. Home
ownership is roughly equal across groups although the second group have slightly
higher housing wealth.

Both groups are similarly long-lived when using ONS life tables to calculate
life expectancy based on gender, age and the year the interview was carried out
in. Subjective life expectancies are also similar across groups, with
individuals expecting to live another 21 years as opposed to the 24 that the ONS
would expect them to live.

@@

\textbf{Maybe add in prob of survival to age 85 or 90 or something}
\textbf{Also add in an indicator for living alone or living with someone else}

Overall, the groups appear to have slightly different wealth profiles with the
treatment group being richer and slightly more likely to have DC pension wealth.
On key demographic characteristics, however, the groups are similar. The
difference in retirement age between the two groups can be mostly explained by
the difference in expected retirement age meaning that individuals have not en
masse decided to delay retirement and avoid annuitisation. Moreover, for the
regression discontinuity models I only use individuals who have a DC pension so
it does not matter that there are more individuals with DC pensions in the
second group.
\begin{landscape}
    \tiny
    \input{tables/r_summary_stats.tex}
    \normalsize
\end{landscape}

As a further test for balance I regress demographic and financial
characteristics of individuals on the treatment dummy. We can then see if the
treatment groups differ on key characteristics such as financial wealth or
retirement age. For a regression discontinuity to be valid we need the treatment
and control groups to be similar along all other characteristics apart from the
treatment variable.

In particular I run:
\begin{equation*}
    Y_{i} = \alpha + \beta PostRef_{i}  + \epsilon_{i}
\end{equation*}

Where $Y_{i}$ are the different demographic and financial characteristics.

\input{tables/cov_balance.tex}

Table \ref{tab:cov_balance} shows these results. As expected, retirement year
and interview year are greater for the treated group. The difference between
expected and real retirement age is 0 which shows that individuals are not
delaying retirement in order to be part of the treatment group although
retirement age is higher post reform. DC pension value is higher, as is
financial wealth but these are both imprecisely estimated. Because of the
differences in these variables I add them to the regression model, following

One threat to validity is manipulation of the running variable, retirement year.
This is probably quite a threat in this setting as individuals could delay
retirement and/or the purchase of an annuity until after the reform. As observed
above in Tables \ref{tab:sum_stats} and \ref{tab:cov_balance}, it does not
appear like individuals are delaying retirement. However, it is possible that
individuals do not buy an annuity at the time of retirement since the law prior
to 2014 only required that they access the pot by age 75. They may decide to
keep their DC pension untouched and live off other income and assets before
accessing it later. Since we do not have data on the exact purchase date of
annuities we cannot track whether annuity purchases were delayed because of
this. However, there is relatively stable annuity demand before the policy
reform. If individuals delayed annuity purchases in expectation of the reform,
so they would not have to buy an annuity at all, we would observe declines in
purchases before the reform was announced. Likewise, if individuals bought an
annuity early in expectation that the end of the compulsory market would cause
prices to go up, we would observe a big spike in annuity purchases prior to
2014.

@@

In this section, I outline the key empirical models I run with both the simulated
consumption data from the lifecycle models and with the real data from ELSA. I
then see which lifecycle model better fits the consumption response that
occurred as a result of the pension reform.

I use a fuzzy regression discontinuity design for which I compare the
consumption of individuals who retired after the policy reform to that of people
who retired just before the reform. The key assumption implicit in regression
discontinuities is that nothing else changes at the time of the jump apart from
the policy of interest, and that the policy occurs without individuals
predicting it and altering their behaviour. As I have argued above, the
demographic information in the data suggests that individuals did not delay
retirement and that there was no delay in annuitisation as shown by the quick
and sudden decline in annuity purchases. Moreover, anecdotal evidence from media
and business sources at the time of the announcement show that there was
surprise the government put forward the reform, with Money Marketing describing the
change as a ``bombshell" \cite{money_marketing_announcement}.

Retirement year is the running variable and individuals are treated if
retirement year is greater than 2014 and less than or equal to 2017. I use
consumption data of individuals up to 2 years into retirement so that the sample
size is larger. So if someone retired in 2015 and had consumption data in 2015
and 2017, I include both values. An individual is considered not treated if they
retire before 2013 and after 2011.

Because of the differences in financial characteristics between the groups, I
add controls for financial wealth, housing wealth, whether an individual owns
their own home, their gender and the size of their state pension. The following
regression equation is used.

@@

Where $Cons_{it}$ is one of five different consumption variables and $X_{it}$ is
a set of controls.

Table \ref{tab:DcOnlyRes} shows the results of this specification on the various
consumption indicators. Column 1 has total monthly consumption on the left hand
side of the regression equation. We can see that being in the treatment group is
associated with £74 more a week in spending overall across all categories of
expenditure. This is not statistically significant using robust standard errors.
Being further into retirement at the time of interview is associated with higher
total consumption whilst those who retired later have lower consumption because
years since the start of retirement is associated with higher total consumption.

Columns 2 through 4 use food consumption as the outcome variable. Interestingly,
food consumption inside the house does not increase as a result of the policy
reform but food consumed away from the house does increase. A shock to income,
caused by no longer needing to annuitise, may cause spending on luxury goods to
increase and for spending on their substitutes, like food in the house relative
to food away from home, to decrease. This income effect may be driving the
results we observe in Table \ref{tab:DcOnlyRes} since food consumption outside the
home increases as does expenditure on clothing -- seen in column 5.

One concern is that the group who retired after 2014 are just generally
different to those who retired before and perhaps consume more outside the home
anyway. The financial characteristic data presented in \ref{tab:sum_stats} would
support this hypothesis as it shows all retirees are richer post reform. To test
for these cohort effects, I run a difference in difference regression to
complement the regression discontinuity just shown. I add individuals who have a
DB pension and those with no private pension into the model and interact the
policy change with dummies for having different types of pension. Specifically, I
run:

This makes the coefficient of interest $\beta_{2}$, which shows us the
difference in the change in consumption between the DC-only group and the base
group that has neither a DB or DC pension. Since rules for DB pensions did not
change under the pensions freedom act, those with a DB pension or no pension can
be seen as a natural counterfactual group to judge the consumption of the DC
group against. Likewise, group with no private pension are unaffected by the
reform.
